% Diffusion Models for TSP — Final Project Report
% Generative Neural Networks for the Sciences, Heidelberg University 2025/26
% NOTE: No university/institution logo on cover!

\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage[margin=2.5cm]{geometry}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{microtype}

% Convenience: highlight TODOs
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}

\title{\textbf{Diffusion Models for Combinatorial Optimization:}\\
       \textbf{Reproducing DIFUSCO and Introducing Flow Matching for TSP}}
\author{
  % TODO: fill in names and Matrikelnummern
  Name1 (Matrikelnummer1)
  \and
  Name2 (Matrikelnummer2)
}
\date{March 2026 \\[0.5em]
  \small Code: \url{https://github.com/sena1818/GNNs} \quad Branch: \texttt{flowmatching}
}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
% ~250 words | Authors: Joint
%
% What problem: TSP is NP-hard; classical solvers scale poorly; ML solvers need new approaches.
% What we do: Reproduce DIFUSCO (NeurIPS 2023) from scratch in PyTorch and extend it
%             with Flow Matching as a novel generative framework.
% How: Three variants share the same Gated GCN backbone — D3PM (discrete diffusion),
%      Gaussian DDPM (continuous diffusion), and Flow Matching (deterministic ODE).
% Results: D3PM achieves 1.90% optimality gap on TSP-50 (vs 0.10% in the original paper
%          with 128k training samples; we use 50k). Gaussian DDPM achieves 6.21%.
%          Flow Matching [RESULT PENDING FM RETRAIN].
% Contribution: First systematic comparison of FM vs DDPM for graph-based CO;
%               full alignment with DIFUSCO's training and decoding pipeline.

The Travelling Salesman Problem (TSP) is a canonical NP-hard combinatorial optimisation
problem with broad practical applications. Recent work has shown that diffusion generative
models can learn to predict high-quality TSP solutions by treating the binary edge adjacency
matrix as a structured object to denoise. In this project we (i) reproduce DIFUSCO
\citep{sun2023difusco} from scratch in plain PyTorch without the original Lightning framework,
and (ii) extend the comparison with Flow Matching \citep{lipman2022flow}, a deterministic
ODE-based generative framework not previously studied for combinatorial optimisation.

All three generative frameworks — discrete D3PM, Gaussian DDPM, and Flow Matching — share an
identical Gated Graph Convolutional Network (Gated GCN) backbone with 12 layers and 256
hidden dimensions, enabling a controlled comparison of the generative modelling paradigm alone.
Models are trained on 50{,}000 TSP-50 instances and decoded with the \texttt{merge\_tours}
greedy-edge-insertion algorithm followed by 2-opt local search.

On TSP-50, D3PM achieves a \textbf{1.90\% optimality gap} (100\% valid tour rate),
Gaussian DDPM achieves \textbf{6.21\%} (100\% valid tour rate after fixing an inference bug),
and Flow Matching achieves \TODO{X.XX\%} after retraining with the correct \(\{-1,+1\}\)
scaling. Our implementation, debugging notes, and experimental results are publicly available
at the repository listed above.
\end{abstract}


% ============================================================
\section{Introduction}
\label{sec:intro}
\textit{[Primary Author: \TODO{Name}]}

The Travelling Salesman Problem asks: given $N$ cities with pairwise distances, find the
shortest closed tour that visits every city exactly once. Despite its simple formulation,
TSP is NP-hard \citep{garey1979computers}, and the best exact solvers (e.g.\ Concorde, LKH)
require exponential time in the worst case, making them impractical for large instances or
time-critical applications.

Neural combinatorial optimisation has emerged as a promising alternative: rather than
searching for a solution at test time, a model learns a solution distribution from data and
produces high-quality solutions in milliseconds. Early approaches used
pointer networks \citep{vinyals2015pointer} or attention-based auto-regressive models
\citep{kool2019attention} to construct tours sequentially. More recently, diffusion
generative models have been applied to the problem in a fundamentally different way:
the binary edge adjacency matrix $\mathbf{A} \in \{0,1\}^{N \times N}$ of the optimal tour
is treated as a structured data object to be generated \citep{sun2023difusco}.

\paragraph{DIFUSCO.}
DIFUSCO \citep{sun2023difusco} introduced this graph diffusion approach (NeurIPS 2023),
demonstrating state-of-the-art results on TSP and MIS. Two variants are proposed: a
\emph{discrete} diffusion model (D3PM \citep{austin2021d3pm}) that corrupts the binary
adjacency matrix by randomly flipping bits, and a \emph{continuous} Gaussian DDPM that
relaxes the matrix to real values and adds Gaussian noise. Both share a Gated GCN encoder
conditioned on node coordinates and the current diffusion timestep.

\paragraph{Motivation for Flow Matching.}
Flow Matching \citep{lipman2022flow} is a recently proposed generative framework that trains
a neural network to predict a \emph{constant velocity field} along straight-line paths from
data to noise. Compared to DDPM, it requires far fewer inference steps (20 vs 50--1000),
has a simpler training objective (MSE on a constant target), and produces more stable
intermediate states. These properties make it an attractive candidate for combinatorial
optimisation, yet no prior work has evaluated it in this setting.

\paragraph{Contributions.}
This project makes the following contributions:
\begin{enumerate}
  \item We reproduce DIFUSCO's D3PM and Gaussian DDPM in plain PyTorch, aligning all
        implementation details (Q-bar transition matrices, Bayes posterior, DDPM stochastic
        inference, \texttt{merge\_tours} decoder) with the official codebase.
  \item We introduce \textbf{Flow Matching as a new generative framework for TSP}, the first
        such comparison in the literature.
  \item We conduct a controlled three-way comparison under identical GNN architecture,
        training data, and decoding pipeline.
  \item We document and fix two non-trivial inference bugs (DDIM accumulation explosion and
        decoder fragility) that caused the Gaussian DDPM valid tour rate to drop to 4\%.
\end{enumerate}

The remainder of the paper is structured as follows. Section~\ref{sec:background} provides
background on TSP, diffusion models, and Flow Matching. Section~\ref{sec:method} describes
our implementation. Section~\ref{sec:experiments} presents experimental results.
Section~\ref{sec:conclusion} concludes.


% ============================================================
\section{Background}
\label{sec:background}
\textit{[Primary Author: \TODO{Name}]}

\subsection{Travelling Salesman Problem}
Given $N$ cities with coordinates $\{\mathbf{c}_i\}_{i=1}^N \subset [0,1]^2$, TSP asks for
a permutation $\pi$ minimising the total tour length:
\begin{equation}
  \min_\pi \sum_{i=1}^N \|\mathbf{c}_{\pi(i)} - \mathbf{c}_{\pi(i+1)}\|_2,
  \quad \pi(N+1) \coloneqq \pi(1).
  \label{eq:tsp}
\end{equation}
The optimal tour corresponds to a binary adjacency matrix
$\mathbf{A}^* \in \{0,1\}^{N \times N}$ with exactly two ones per row and column, forming a
single Hamiltonian cycle. We measure solution quality as the \emph{optimality gap}:
\begin{equation}
  \text{gap} = \frac{\text{cost}(\hat\pi) - \text{cost}(\pi^*)}{\text{cost}(\pi^*)} \times 100\%.
\end{equation}

\subsection{Denoising Diffusion Probabilistic Models (DDPM)}
DDPM \citep{ho2020ddpm} defines a forward process that gradually adds Gaussian noise to data
$\mathbf{x}_0$ over $T$ steps:
\begin{equation}
  q(\mathbf{x}_t \mid \mathbf{x}_0)
  = \mathcal{N}\!\left(\sqrt{\bar\alpha_t}\,\mathbf{x}_0,\;(1-\bar\alpha_t)\mathbf{I}\right),
  \quad \bar\alpha_t = \prod_{s=1}^t (1-\beta_s),
  \label{eq:ddpm_forward}
\end{equation}
where $\{\beta_t\}$ is a noise schedule. A neural network $\epsilon_\theta$ is trained to
predict the noise:
\begin{equation}
  \mathcal{L}_\text{DDPM}
  = \mathbb{E}_{t,\mathbf{x}_0,\boldsymbol\epsilon}
    \left[\|\epsilon_\theta(\mathbf{x}_t, t) - \boldsymbol\epsilon\|^2\right].
  \label{eq:ddpm_loss}
\end{equation}
Inference proceeds by iterating the DDPM posterior:
\begin{equation}
  \mathbf{x}_{t-1}
  = \frac{1}{\sqrt{\alpha_t}}\!\left(\mathbf{x}_t - \frac{1-\alpha_t}{\sqrt{1-\bar\alpha_t}}\,
    \epsilon_\theta(\mathbf{x}_t,t)\right) + \sqrt{\tilde\beta_t}\,\mathbf{z},
  \quad \mathbf{z}\sim\mathcal{N}(\mathbf{0},\mathbf{I}),
  \label{eq:ddpm_posterior}
\end{equation}
where $\tilde\beta_t = \beta_t(1-\bar\alpha_{t-1})/(1-\bar\alpha_t)$.
The optional deterministic DDIM sampler \citep{song2020ddim} removes the noise term $\mathbf{z}$
but is more susceptible to error accumulation over many steps (see Section~\ref{sec:bugs}).

\subsection{D3PM: Discrete Diffusion}
\label{sec:d3pm}
For binary data, D3PM \citep{austin2021d3pm} defines a discrete Markov chain using
$2\times 2$ transition matrices:
\begin{equation}
  \mathbf{Q}_t = (1-\beta_t)\mathbf{I} + \tfrac{\beta_t}{2}\mathbf{1}\mathbf{1}^\top,
  \quad \bar{\mathbf{Q}}_t = \mathbf{Q}_1\cdots\mathbf{Q}_t.
\end{equation}
Forward sampling: $q(\mathbf{x}_t|\mathbf{x}_0)
= \text{Cat}(\mathbf{x}_0\bar{\mathbf{Q}}_t)$.
The Bayes posterior $q(\mathbf{x}_{t-1}|\mathbf{x}_t,\hat{\mathbf{x}}_0)$ is computed
analytically and used for inference (see DIFUSCO \citep{sun2023difusco} Appendix for
the full formula). Crucially, all intermediate states $\mathbf{x}_t$ remain binary
graphs, preserving the adjacency structure throughout the diffusion trajectory.

\subsection{Discrete vs.\ Continuous Diffusion for Graphs}
\label{sec:disc_vs_cont}
Discrete diffusion preserves binary graph structure at every timestep, giving the GNN
a natural message-passing domain throughout the trajectory. Continuous Gaussian diffusion
relaxes entries to $\mathbb{R}$, producing intermediate states with no natural graph
interpretation. DIFUSCO reports $\sim\!0.10$--$0.17\%$ gap with D3PM versus
$\sim\!0.25$--$0.30\%$ with Gaussian DDPM on TSP-50, a consistent gap across decoding
strategies. This motivates the question: can a better continuous-space generative framework
narrow this gap?

\subsection{Flow Matching}
\label{sec:fm_background}
Flow Matching \citep{lipman2022flow} trains a neural ODE by regressing on the velocity of
straight-line paths between data and noise distributions. Given
$\mathbf{x}_0 \sim p_\text{data}$ and $\boldsymbol\epsilon \sim \mathcal{N}(\mathbf{0},\mathbf{I})$:
\begin{align}
  \mathbf{x}_t &= (1-t)\,\mathbf{x}_0 + t\,\boldsymbol\epsilon,
    \quad t\sim\mathcal{U}[0,1], \label{eq:fm_interp} \\
  \mathbf{v}^*(\mathbf{x}_t, t) &= \boldsymbol\epsilon - \mathbf{x}_0. \label{eq:fm_vtarget}
\end{align}
The velocity target is \emph{constant} along each path, leading to the simple MSE loss:
\begin{equation}
  \mathcal{L}_\text{FM}
  = \mathbb{E}_{t,\mathbf{x}_0,\boldsymbol\epsilon}
    \left[\|v_\theta(\mathbf{x}_t,t) - (\boldsymbol\epsilon - \mathbf{x}_0)\|^2\right].
  \label{eq:fm_loss}
\end{equation}
Inference integrates the ODE $\dot{\mathbf{x}} = v_\theta(\mathbf{x},t)$ from $t=1$
(noise) to $t=0$ (data) with $K$ Euler steps, requiring far fewer function evaluations
than DDPM.

\subsection{Related Work}
\TODO{Brief survey: DIFUSCO \citep{sun2023difusco}, T2T-CO \citep{chen2024t2tco},
Attention Model \citep{kool2019attention}, LKH \citep{helsgott2000lkh}, POMO.
2--3 sentences each.}


% ============================================================
\section{Method}
\label{sec:method}
\textit{[Primary Author: Joint]}

\subsection{Architecture Overview}
All three generative frameworks share an identical GNN encoder
$f_\theta(\mathbf{C}, \mathbf{x}_t, t)$ that maps city coordinates $\mathbf{C}$, the
current noisy adjacency state $\mathbf{x}_t$, and timestep $t$ to either a velocity field
(FM), a noise prediction (Gaussian DDPM), or class logits (D3PM). The shared backbone
enables a controlled comparison of the generative paradigm alone.

\TODO{Insert architecture diagram: GNN Encoder block diagram showing node/edge embeddings,
time injection, and output head.}

\subsection{Gated GCN Encoder}
\label{sec:gnn}
We use a 12-layer Gated GCN \citep{bresson2017gatedgcn} with hidden dimension 256,
matching DIFUSCO's default configuration. Each layer performs:
\begin{align}
  \mathbf{g}_{ij} &= \sigma\!\left(
    \mathbf{A}\mathbf{h}_i + \mathbf{B}\mathbf{h}_j + \mathbf{C}\mathbf{e}_{ij}
  \right), \label{eq:gate} \\
  \mathbf{h}_i' &= \text{ReLU}\!\left(\mathbf{U}\mathbf{h}_i
    + \textstyle\sum_j \mathbf{g}_{ij} \odot \mathbf{V}\mathbf{h}_j\right), \\
  \mathbf{e}_{ij}' &= \text{ReLU}\!\left(\mathbf{A}\mathbf{h}_i + \mathbf{B}\mathbf{h}_j
    + \mathbf{C}\mathbf{e}_{ij}\right),
\end{align}
where $\sigma$ is the sigmoid function and $\odot$ is element-wise multiplication.
External residual connections are applied after each layer. City coordinates are encoded
via sinusoidal position embeddings; the diffusion timestep $t$ is encoded as a sinusoidal
scalar embedding injected into every edge feature map.

The output head consists of GroupNorm $\to$ ReLU $\to$ Conv2d$(d, C_\text{out}, 1)$,
with $C_\text{out}=2$ for D3PM (class logits) and $C_\text{out}=1$ for Gaussian DDPM and FM
(scalar prediction).

\subsection{D3PM Training and Inference}
\label{sec:d3pm_method}
\paragraph{Training.}
Given ground-truth $\mathbf{A}_0 \in \{0,1\}^{N\times N}$, we sample
$t \sim \text{Uniform}\{1,\ldots,T\}$, apply the forward process
$\mathbf{x}_t \sim \text{Cat}(\mathbf{A}_0^{1\text{-hot}} \bar{\mathbf{Q}}_t)$,
scale to $\{-1,+1\}$ and add 5\% jitter (following DIFUSCO), and train with cross-entropy:
\begin{equation}
  \mathcal{L}_\text{D3PM} = \text{CrossEntropy}\!\left(f_\theta(\mathbf{C}, \mathbf{x}_t, t),\;
    \mathbf{A}_0\right).
\end{equation}

\paragraph{Inference.}
Starting from $\mathbf{x}_T \in \{0,1\}^{N\times N}$ (random Bernoulli), we apply 50 steps
of the categorical posterior $q(\mathbf{x}_{t-1}|\mathbf{x}_t, \hat{\mathbf{x}}_0)$
computed via Bayes rule with the cumulative transition matrices $\bar{\mathbf{Q}}_t$.
The final heatmap is the model's softmax probability for the edge class:
$\hat{\mathbf{P}} = \text{softmax}(f_\theta)[...,1] \in [0,1]^{N\times N}$.

\subsection{Gaussian DDPM Training and Inference}
\label{sec:gaussian_method}
\paragraph{Training.}
The adjacency matrix is scaled to $\{-1,+1\}$ (plus 5\% jitter) before noising.
A linear beta schedule with $T=1000$ steps is used. The GNN predicts the noise $\epsilon$
and is trained with MSE (Equation~\ref{eq:ddpm_loss}).

\paragraph{Inference.}
We apply 50 steps of the DDPM \emph{stochastic} posterior
(Equation~\ref{eq:ddpm_posterior}), matching DIFUSCO's default. The deterministic DDIM
sampler is available via a flag but causes error accumulation over 50 steps
(Section~\ref{sec:bugs}). The final heatmap is obtained by $\hat{\mathbf{P}} = \mathbf{x}_0 \cdot 0.5 + 0.5$.

\subsection{Flow Matching for TSP}
\label{sec:fm_method}
\paragraph{Training.}
We apply the \(\{-1,+1\}\) scaling $\mathbf{A}_\text{scaled} = 2\mathbf{A}_0 - 1$
to improve the separability of the velocity field: edge entries target velocity
$v^* = \epsilon - 1$ (mean $-1$) while non-edge entries target $v^* = \epsilon + 1$
(mean $+1$), giving a mean separation of 2 compared to 1 in the raw $\{0,1\}$ space.
The FM loss (Equation~\ref{eq:fm_loss}) is computed on this scaled space.

\paragraph{Inference.}
Starting from $\mathbf{x}_1 \sim \mathcal{N}(\mathbf{0},\mathbf{I})$, we integrate
with $K=20$ Euler steps:
\begin{equation}
  \mathbf{x}_{t-\Delta t} = \mathbf{x}_t - \Delta t \cdot v_\theta(\mathbf{x}_t, t, \mathbf{C}),
  \quad \Delta t = 1/K.
\end{equation}
The output is mapped back: $\hat{\mathbf{P}} = \mathbf{x}_0 \cdot 0.5 + 0.5$,
clipped to $[0,1]$ and symmetrised.

\subsection{Tour Decoding}
\label{sec:decoding}
All methods produce a symmetric heatmap $\hat{\mathbf{P}} \in [0,1]^{N\times N}$.
We decode it to a valid Hamiltonian cycle using \texttt{merge\_tours}
\citep{sun2023difusco}, which greedily inserts edges sorted by score
$s_{ij} = \hat{P}_{ij} / d_{ij}$ (heatmap probability over Euclidean distance),
using a union-find data structure to prevent degree-$>$2 violations and premature
cycle closure. This guarantees a valid tour even from noisy heatmaps.
The resulting tour is refined with 2-opt local search.

\subsection{Hyperparameters}

\begin{table}[h]
\centering
\caption{Shared hyperparameters for all three models.}
\label{tab:hyperparams}
\begin{tabular}{lll}
\toprule
Parameter & Value & Note \\
\midrule
GNN layers & 12 & DIFUSCO default \\
Hidden dimension & 256 & DIFUSCO default \\
Batch size & 64 & RTX 4090 24\,GB \\
Learning rate & $2\times10^{-4}$ & AdamW \\
Weight decay & $10^{-4}$ & AdamW \\
Epochs & 50 & \\
LR schedule & CosineAnnealing & \\
EMA decay & 0.999 & \\
Gradient clip & 1.0 & max norm \\
Diffusion steps $T$ & 1000 & D3PM \& Gaussian \\
Inference steps & 50 & D3PM \& Gaussian \\
Inference steps & 20 & Flow Matching \\
Training instances & 50{,}000 & TSP-50 \\
Test instances & 1{,}000 & TSP-50 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Deviations from Original Plan}
\label{sec:deviations}
% REQUIRED by course guidelines

Several deviations from our initial plan occurred during implementation:

\paragraph{Decoder.}
The original plan used a simple greedy decoder. After discovering a 4\% valid tour rate
for Gaussian DDPM, we switched to \texttt{merge\_tours} (the official DIFUSCO decoder)
for all models, which guarantees valid Hamiltonian cycles.

\paragraph{Gaussian inference.}
We initially implemented only DDIM (deterministic) inference for Gaussian DDPM.
Comparison with the official codebase revealed that DIFUSCO defaults to DDPM
\emph{stochastic} inference; DDIM caused error accumulation over 50 steps,
driving the valid tour rate to 4\%. We added the stochastic posterior and made it default.

\paragraph{FM target space.}
The FM model was initially trained with raw $\{0,1\}$ targets. We later scaled to
$\{-1,+1\}$ (matching Gaussian DDPM pre-processing) to improve velocity field
separability, requiring a full retraining.

\paragraph{Training data.}
The original plan targeted 5{,}000 training instances (matching DIFUSCO's small-scale
setting). We generated and used 50{,}000 instances throughout.


% ============================================================
\section{Experiments}
\label{sec:experiments}
\textit{[Primary Author: Joint]}

\subsection{Setup}
All models are trained on an NVIDIA RTX 4090 (24\,GB) for approximately 2 hours each.
The test set consists of 1{,}000 TSP-50 instances. Optimality gaps are computed against
LKH solutions provided in the dataset. All three models use identical training data,
architecture, and decoding pipeline.

\subsection{Training Convergence}
\TODO{Insert Figure: training and validation loss curves for all three models over 50 epochs.
Use history.json files. D3PM converges from 0.0624 to 0.0122 (train) / 0.0445 to 0.0139 (val).
Continuous DDPM: 0.0513 to 0.0052 / 0.0198 to 0.0052.
FM: 0.1265 to 0.0081 / 0.0679 to 0.0081.
Note: FM starts higher because MSE on velocity has a different loss scale than cross-entropy.}

\subsection{Main Results: Three-Way Comparison}
\label{sec:three_way}
Table~\ref{tab:main} presents the main comparison. All gaps are computed on the 1{,}000
TSP-50 test instances using the \texttt{merge\_tours} decoder followed by 2-opt.

\begin{table}[h]
\centering
\caption{Optimality gap on TSP-50 (1{,}000 test instances). Lower is better.
Decoder: \texttt{merge\_tours} + 2-opt for all.}
\label{tab:main}
\begin{tabular}{lcccccc}
\toprule
Method & Type & Steps & Avg Gap (\%) & Std (\%) & Valid & Time (ms) \\
\midrule
D3PM (discrete) & $\{0,1\}$ & 50 & \textbf{1.90} & 2.02 & 100\% & 403 \\
Gaussian DDPM   & $\mathbb{R}$ & 50 & 6.21 & 3.56 & 100\% & 334 \\
Flow Matching   & $\mathbb{R}$ (ODE) & 20 & \TODO{X.XX} & \TODO{X.XX} & \TODO{XX\%} & \TODO{XXX} \\
\midrule
\multicolumn{7}{l}{\textit{Reference (DIFUSCO official, 128k train samples)}} \\
D3PM            & $\{0,1\}$ & 50 & 0.10--0.17 & — & 100\% & — \\
Gaussian DDPM   & $\mathbb{R}$ & 50 & 0.25--0.30 & — & 100\% & — \\
\bottomrule
\end{tabular}
\end{table}

The gap between our results and DIFUSCO's original is primarily attributable to the
training data size (50k vs 128k instances). Our D3PM gap of 1.90\% remains competitive
for a from-scratch re-implementation at reduced data scale.
D3PM outperforms Gaussian DDPM, consistent with DIFUSCO's findings that discrete
diffusion preserves the binary graph structure more faithfully during training.

\subsection{Inference Bug Analysis}
\label{sec:bugs}

\paragraph{DDIM accumulation (Gaussian DDPM).}
When using the deterministic DDIM sampler for Gaussian DDPM inference, the valid tour rate
collapsed to 4\% despite the model converging well (val loss 0.0052). Analysis showed that
prediction errors in $\epsilon_\theta$ accumulate deterministically over 50 steps, driving
$\mathbf{x}_t$ to extreme negative values. The heatmap $\hat{\mathbf{P}} = \mathbf{x}_0\cdot 0.5+0.5$
then saturates near 0, causing the decoder to fail. Switching to the DDPM stochastic posterior
(Equation~\ref{eq:ddpm_posterior}) — which injects fresh noise $\mathbf{z}$ at each step,
disrupting error accumulation — restored the valid rate to 100\%.

\paragraph{Decoder fragility (all models).}
The initial greedy decoder (following the highest heatmap edge from node 0) failed
catastrophically when heatmap rows were near-uniform. The \texttt{merge\_tours} decoder,
which sorts \emph{all} edges globally and uses union-find to enforce degree and cycle
constraints, is robust to noisy heatmaps by construction.

\subsection{Flow Matching Analysis}
\label{sec:fm_analysis}
\TODO{After FM retraining completes, fill in: final gap, comparison vs DDPM,
effect of {-1,+1} scaling vs raw {0,1}.}

\TODO{Optional: Table showing FM gap at different inference steps K = 10, 20, 50 to show
the speed--quality tradeoff.}

\subsection{Diffusion Process Visualisation}
\TODO{Insert heatmap figure showing intermediate states at t = 1.0, 0.75, 0.5, 0.25, 0.0
for one TSP-50 instance. Should show progression from uniform noise to sparse edge structure.}

\TODO{Insert tour comparison figure: LKH optimal tour (left) vs model decoded tour (right)
for a representative instance with low, medium, and high gap.}


% ============================================================
\section{Conclusion}
\label{sec:conclusion}
\textit{[Primary Author: Joint]}

We reproduced DIFUSCO's D3PM and Gaussian DDPM models from scratch in plain PyTorch,
achieving 1.90\% and 6.21\% optimality gap on TSP-50 respectively, and extended the
comparison with Flow Matching as a new generative framework for combinatorial optimisation.

Our reproduction confirms the key finding of DIFUSCO: discrete diffusion (D3PM) consistently
outperforms continuous Gaussian DDPM for graph-structured combinatorial optimisation.
The binary-preserving transition matrices of D3PM keep the adjacency structure interpretable
throughout the diffusion trajectory, providing a cleaner learning signal for the GNN.

Flow Matching offers a distinct tradeoff: it requires only 20 inference steps (vs 50 for
DDPM) and has a simpler training objective, but operates in a continuous space that does
not naturally preserve binary graph structure. \TODO{Whether this tradeoff is favourable
will be discussed after final results.}

\paragraph{Limitations.}
Our training set (50k instances) is smaller than DIFUSCO's (128k), accounting for
the performance gap. We use a Python 2-opt implementation rather than DIFUSCO's GPU-accelerated
version, which slightly inflates inference times. Parallel and sequential sampling are not implemented.

\paragraph{Future work.}
Promising directions include: (i) second-order ODE solvers (Heun's method) for FM to
improve accuracy with fewer steps; (ii) mixed discrete-continuous diffusion paths that
preserve binary structure while enabling FM-style training; (iii) scaling to TSP-100 and
beyond with sparse graph representations.


% ============================================================
\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem[Austin et al.(2021)]{austin2021d3pm}
Austin, J., Johnson, D.~D., Ho, J., Tarlow, D., \& van den Berg, R. (2021).
Structured Denoising Diffusion Models in Discrete State-Spaces.
\textit{NeurIPS 2021}.

\bibitem[Bresson \& Laurent(2017)]{bresson2017gatedgcn}
Bresson, X., \& Laurent, T. (2017).
Residual Gated Graph ConvNets.
\textit{arXiv:1711.07553}.

\bibitem[Chen et al.(2024)]{chen2024t2tco}
Chen, T. et al. (2024).
Fast T2T: Optimization Consistency Speeds Up Diffusion-Based Training-to-Testing
for Combinatorial Optimization.
\textit{NeurIPS 2024}.

\bibitem[Garey \& Johnson(1979)]{garey1979computers}
Garey, M.~R., \& Johnson, D.~S. (1979).
\textit{Computers and Intractability: A Guide to the Theory of NP-Completeness}.
W.~H. Freeman.

\bibitem[Ho et al.(2020)]{ho2020ddpm}
Ho, J., Jain, A., \& Abbeel, P. (2020).
Denoising Diffusion Probabilistic Models.
\textit{NeurIPS 2020}.

\bibitem[Kool et al.(2019)]{kool2019attention}
Kool, W., van Hoof, H., \& Welling, M. (2019).
Attention, Learn to Solve Routing Problems!
\textit{ICLR 2019}.

\bibitem[Lipman et al.(2022)]{lipman2022flow}
Lipman, Y., Chen, R. T.~Q., Ben-Hamu, H., Nickel, M., \& Le, M. (2022).
Flow Matching for Generative Modeling.
\textit{ICLR 2023}.

\bibitem[Song et al.(2020)]{song2020ddim}
Song, J., Meng, C., \& Ermon, S. (2020).
Denoising Diffusion Implicit Models.
\textit{ICLR 2021}.

\bibitem[Sun et al.(2023)]{sun2023difusco}
Sun, Z., Yang, Y., \& Zheng, J. (2023).
DIFUSCO: Graph-based Diffusion Solvers for Combinatorial Optimization.
\textit{NeurIPS 2023}.

\bibitem[Veli\v{c}kovi\'{c} et al.(2018)]{velickovic2018gat}
Veli\v{c}kovi\'{c}, P., Cucurull, G., Casanova, A., Romero, A., Li\`{o}, P., \& Bengio, Y. (2018).
Graph Attention Networks.
\textit{ICLR 2018}.

\bibitem[Vinyals et al.(2015)]{vinyals2015pointer}
Vinyals, O., Fortunato, M., \& Jaitly, N. (2015).
Pointer Networks.
\textit{NeurIPS 2015}.

\end{thebibliography}

\end{document}
